---
title: "Introduction to R"
subtitle: "Import and Clean Data and Make Tables in R"
output: html_document
---

## How to follow along and participate

```{r chunk1}
library(package = "here")
library(package = "tidyverse")
library(package = "haven")
library(package = "table1")
```

## Importing data into R

```{r chunk2}
# use the here() function so that R looks in the same folder you are in
here()

# use read_spss function to import the spss file 
globalData2019 <- read_spss(file = "Pew Research Center Global Attitudes Spring 2019 Dataset WEB.sav")
```

## Selecting variables and filtering observations

```{r chunk3}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) 

```

## Examining the smaller data set

```{r chunk4}
summary(object = globalData2019clean)
```

## Changing data types 1

```{r chunk5}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States'))
```

## Checking your work 1

```{r chunk6}
summary(object = globalData2019clean)
```

## Make a graph

```{r chunk7}
globalData2019clean %>% 
  ggplot(aes(x = country)) +
  geom_bar()
```

## Make a fancier graph

```{r chunk8}
globalData2019clean %>% 
  ggplot(aes(x = country)) +
  geom_bar(fill = '#3264ce') +
  theme_minimal() +
  labs(y = "Number of participants",
       x = "Country of residence",
       title = "Survey participation by country of residence")
  
```

## Break 

* Some fun things to try on your break:

    + Choose a new color from colorbrewer2.org and use it to change your bar colors 
    
    + Add the text `base_size = 16` inside the parentheses in the `theme_minimal()` layer and see what happens when you run the code 
    
    + Add the text `base_family = 'serif'` inside the parentheses in the `theme_minimal()` layer and see what happens when you run the code 
        
        + You can keep the `base_size` argument and add `base_family` like `base_size = 16, base_family = 'serif'` or replace `base_size` with `base_family`

## Reminder: Changing data types 1

```{r chunk9}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States'))
```

## Changing data types 2

```{r chunk10}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States')) %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) 
```

## Checking your work 2

```{r chunk11}
summary(object = globalData2019clean)
```

## Changing data types 3

```{r chunk12}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States')) %>%
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_))
```


## Checking your work 3

```{r chunk13}
summary(object = globalData2019clean)
```

## Recoding data: You Try It!

* Look in the codebook to find the variable for ECON_SIT (note: it will be lower case in the codebook)

* Add recoding of ECON_SIT to the code below and run the code

* Run the summary() code in the next code chunk to check your work

```{r chunk14}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States')) %>%
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) 
```


```{r chunk15}
# check your work
summary(object = globalData2019clean)
```


## Try Some More! 

* Recode the last two categorical variables, `BELIEVE_GOD` and `GLOBAL_COMMUNITY` to have appropriate category labels and `NA` 

* Use the codebook to find the labels you need 

* Add to the existing code: 

```{r chunk16}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States')) %>%
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_))
```

```{r chunk17}
# check your work
summary(object = globalData2019clean)
```


## Final recoding for age

```{r chunk18}
globalData2019clean <- globalData2019 %>%
  filter(country %in% c(16, 21, 26, 34)) %>% 
  select(country, SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>%
  mutate(AGE = na_if(AGE, 98)) %>% 
  mutate(country = recode_factor(country,
                             '16' = 'Kenya',
                             '21' = 'Nigeria',
                             '26' = 'South Africa',
                             '34' = 'United States')) %>%
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_)) %>% 
  mutate(BELIEVE_GOD = recode_factor(BELIEVE_GOD,
                                     '1' = 'It is not necessary to believe in God in order to be moral and have good values',
                                     '2' = 'It is necessary to believe in God in order to be moral and have good values',
                                     '8' = NA_character_,
                                     '9' = NA_character_)) %>% 
  mutate(GLOBAL_COMMUNITY = recode_factor(GLOBAL_COMMUNITY,
                                          '1' = 'Part of a global community that works together to solve problems',
                                          '2' = 'Independent nations that compete with other countries and pursue their own interests',
                                          '3' = 'Both',
                                          '4' = 'Neither',
                                          '8' = NA_character_,
                                          '9' = NA_character_)) 

```


## Basic table

* We can start with a simple table with the `table1` package that shows the `SEX`, `AGE`, and `BETTER_GENDER` variables with their descriptive statistics

```{r chunk19}
table1(~ SEX + AGE + BETTER_GENDER,
       data = globalData2019clean)
```

## Basic table with labels

* Add labels for the variables and add the `BELIEVE_GOD` variable

```{r chunk20}
label(globalData2019clean$SEX) <- "Sex"
label(globalData2019clean$AGE) <- "Age (years)" 
label(globalData2019clean$BETTER_GENDER) <- "Who has a better life in this country"
label(globalData2019clean$BELIEVE_GOD) <- "Which comes closest to your opinion"
```


## Basic table with labels

```{r chunk21}
table1(~ SEX + AGE + BETTER_GENDER + BELIEVE_GOD,
       data = globalData2019clean)
```

## Basic table with labels and grouped by sex

```{r chunk22}
table1(~ AGE + BETTER_GENDER + BELIEVE_GOD | SEX,
       data = globalData2019clean)
```

## Basic table with labels, groups, and total

```{r chunk23}
table1(~ AGE + BETTER_GENDER + BELIEVE_GOD | SEX,
       data = globalData2019clean, overall = "Total")
```


## You Try It!

* Copy and paste the label and table code from above into the empty R chunk

* Add labels to the `COUNTRY_SATIS` and `ECON_SIT` variables (check the codebook to find these)

* Add the `COUNTRY_SATIS` and `ECON_SIT` variables to the table 

* Run the code to make sure it works

```{r chunk24}

```


## The end

* Email me at [harrisj@wustl.edu](mailto:harrisj@wustl.edu)

* Connect with me on Twitter [@jenineharris](https://twitter.com/jenineharris)

